---
title: "Bellabeat Case Study"
author: "Vaishnavi Devadig"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---


```{r}
library(tidyverse)
dailyActivity_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
dailyCalories_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
dailyIntensities_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
dailySteps_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")
heartrate_seconds_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")
hourlyCalories_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
hourlyIntensities_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv")
hourlySteps_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv")
minuteCaloriesNarrow_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteCaloriesNarrow_merged.csv")
minuteCaloriesWide_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteCaloriesWide_merged.csv")
minuteIntensitiesNarrow_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteIntensitiesNarrow_merged.csv")
minuteIntensitiesWide_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteIntensitiesWide_merged.csv")
minuteMETsNarrow_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteMETsNarrow_merged.csv")
minuteSleep_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteSleep_merged.csv")
minuteStepsNarrow_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteStepsNarrow_merged.csv")
minuteStepsWide_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/minuteStepsWide_merged.csv")
sleepDay_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weightLogInfo_merged = read_csv("/Users/vaishnavishankardevadig/Downloads/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
```




# Checking for the dataset dailyActivity_merged, converting the ActivityDate column from Character to Date type.

```{r}
head(dailyActivity_merged)
```

```{r}
library(lubridate)
dailyActivity_merged$ActivityDate = as.Date(dailyActivity_merged$ActivityDate, format='%m/%d/%Y')
head(dailyActivity_merged)
```

# Checking for null values

```{r}
nulldam = colSums(is.na(dailyActivity_merged))
nulldam
```


# Changing the format for date in the datasets for daily calories, intensities and steps

```{r}
dailyCalories_merged$ActivityDay = as.Date(dailyCalories_merged$ActivityDay, format='%m/%d/%Y')
dailyIntensities_merged$ActivityDay = as.Date(dailyIntensities_merged$ActivityDay, format='%m/%d/%Y')
dailySteps_merged$ActivityDay = as.Date(dailySteps_merged$ActivityDay, format='%m/%d/%Y')
head(dailyCalories_merged)
head(dailyIntensities_merged)
head(dailySteps_merged)
```

# Checking for null values

```{r}
ndcm = colSums(is.na(dailyCalories_merged))
ndim = colSums(is.na(dailyIntensities_merged))
ndsm = colSums(is.na(dailySteps_merged))
print("dailyCalories_merged")
ndcm
cat("\n")
print("dailyIntensities_merged")
ndim
cat("\n")
print("dailySteps_merged")
ndsm
cat("\n")
```


# Converting the respective date and time columns to POSIX objects for the remaining datasets.

```{r}
minuteCaloriesNarrow_merged$ActivityMinute <- as.POSIXct(minuteCaloriesNarrow_merged$ActivityMinute, format = "%m/%d/%Y %I:%M:%S %p")
minuteCaloriesWide_merged$ActivityHour <- as.POSIXct(minuteCaloriesWide_merged$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
minuteIntensitiesNarrow_merged$ActivityMinute <- as.POSIXct(minuteIntensitiesNarrow_merged$ActivityMinute, format = "%m/%d/%Y %I:%M:%S %p")
minuteIntensitiesWide_merged$ActivityHour <- as.POSIXct(minuteIntensitiesWide_merged$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
minuteMETsNarrow_merged$ActivityMinute <- as.POSIXct(minuteMETsNarrow_merged$ActivityMinute, format = "%m/%d/%Y %I:%M:%S %p")
hourlyCalories_merged$ActivityHour <- as.POSIXct(hourlyCalories_merged$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
hourlyIntensities_merged$ActivityHour <- as.POSIXct(hourlyIntensities_merged$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
heartrate_seconds_merged$Time = as.POSIXct(heartrate_seconds_merged$Time, format='%m/%d/%Y %H:%M:%S')
minuteSleep_merged$date = as.POSIXct(minuteSleep_merged$date, format='%m/%d/%Y %H:%M:%S')
sleepDay_merged$SleepDay = as.POSIXct(sleepDay_merged$SleepDay, format='%m/%d/%Y %H:%M:%S')
weightLogInfo_merged$Date = as.POSIXct(weightLogInfo_merged$Date, format='%m/%d/%Y %H:%M:%S')
hourlySteps_merged$ActivityHour <- as.POSIXct(hourlySteps_merged$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
head(minuteCaloriesNarrow_merged)
head(minuteCaloriesWide_merged)
head(minuteIntensitiesNarrow_merged)
head(minuteIntensitiesWide_merged)
head(minuteMETsNarrow_merged)
head(hourlyCalories_merged)
head(hourlyIntensities_merged)
head(hourlySteps_merged)
head(heartrate_seconds_merged)
head(minuteSleep_merged)
head(sleepDay_merged)
head(weightLogInfo_merged)

```


#Checking for null values

```{r}
colSums(is.na(hourlyCalories_merged))
colSums(is.na(hourlyIntensities_merged))
colSums(is.na(hourlySteps_merged))
colSums(is.na(minuteCaloriesNarrow_merged))
colSums(is.na(minuteCaloriesWide_merged))
colSums(is.na(minuteIntensitiesNarrow_merged))
colSums(is.na(minuteIntensitiesWide_merged))
colSums(is.na(minuteMETsNarrow_merged))
colSums(is.na(heartrate_seconds_merged))
colSums(is.na(minuteSleep_merged))
colSums(is.na(sleepDay_merged))
colSums(is.na(weightLogInfo_merged))
```


#Summary of a few datasets in order to identify general trends

```{r}
summary(dailyActivity_merged)
```

### As per studies, on an average, an adult woman should burn about 2000 calories a day, while an adult man should be burning about 2500 calories a day.

### The 'Calories' column shows the mean value to be 2304, which is above average for adult women but below average for adult male.

### Sedentary Minutes shows the mean value to be 991.2, which is about 16.5 hours. This value is extremely high and has scope to be reduced.


```{r}
summary(sleepDay_merged)
```
### The mean total minutes asleep is 6.99 ~ 7 hours, which is also slightly below the accepted average of 8 hours. However, this is an acceptable number.


```{r}
summary(heartrate_seconds_merged)
```

### Changing the various column names for time and date as ActivityDate to facilitate merging

```{r}
colnames(sleepDay_merged)[colnames(sleepDay_merged) == "SleepDay"] <- "ActivityDate"
colnames(heartrate_seconds_merged)[colnames(heartrate_seconds_merged) == "Time"] <- "ActivityDate"
colnames(weightLogInfo_merged)[colnames(weightLogInfo_merged) == "Date"] <- "ActivityDate"
head(dailyActivity_merged)
head(sleepDay_merged)
head(weightLogInfo_merged)
head(heartrate_seconds_merged)
```

### Separating date and time columns 

```{r}
library(tidyverse)

sleepnew_df = separate(sleepDay_merged, ActivityDate, into = c("ActivityDate", "Time"), sep = " ")
weightnew_df = separate(weightLogInfo_merged, ActivityDate, into = c("ActivityDate", "Time"), sep = " ")
heartnew_df = separate(heartrate_seconds_merged, ActivityDate, into = c("ActivityDate", "Time"), sep = " ")
stepsnew_df = separate(hourlySteps_merged, ActivityHour, into = c("ActivityDate", "Time"), sep = " ")
head(sleepnew_df)
head(weightnew_df)
head(heartnew_df)
head(stepsnew_df)
```


### Merging the Activity and Sleep datasets

```{r}
final = merge(dailyActivity_merged, sleepnew_df, by = c('Id', 'ActivityDate'))
final
```


### Creating a new dataframe for mean heart beat rate

```{r}
mean_heart_rate = heartnew_df %>%
  group_by(Id, ActivityDate) %>%
  summarise(Mean_Heartrate = mean(Value))

mean_heart_rate
```

### Merging the new data frame with final

```{r}
final = merge(final, mean_heart_rate, by = c('Id', 'ActivityDate'))
final
```

### Creating data frames for mean steps and calories burnt

```{r}
daybasedsteps = dailyActivity_merged %>%
  group_by(ActivityDate) %>%
  drop_na() %>%
  summarise(Mean_Steps = mean(TotalSteps), Mean_Calories_Burnt = mean(Calories))

daybasedsteps
```

### Combining Heart rate and Weight log info datasets

```{r}
heartandweight = merge(weightnew_df, heartnew_df, by = c('Id', 'ActivityDate', 'Time'))
heartandweight
```
### Not a good enough dataset to perform analysis.


# PLOTS

```{r}
ggplot(daybasedsteps, mapping=aes(x=ActivityDate, y=Mean_Steps))+geom_point(color="lightcoral")+geom_smooth(method="loess", color="pink4")+ labs(title="Average Steps vs. Date")
```

### The intention behind this plot was to identify if we could find weekday-weekend patterns with respect to the number of steps walked in a day. 

### However, there is no specific pattern that can be deduced from the plot.




```{r}
ggplot(dailyActivity_merged, mapping=aes(x=TotalSteps, y=Calories))+geom_point(color="seagreen")+geom_smooth(color="royalblue4")+ labs(title="Calories Burnt vs. Total Steps")
```

### Here, it is clear that with the higher number of steps, the subject also loses more calories. Hence, there is a positive relationship. 

### As a suggestion to Bellabeat, to faciliate more activity amongst the users, a rewards/points program could be introduced based on thr total steps count per day, rather than just relying on a default appreciation message.




```{r}
stepsplot = stepsnew_df %>%
  group_by(Time) %>%
  drop_na() %>%
  summarise(Mean_Steps_By_Time = mean(StepTotal))
ggplot(stepsplot, aes(x=Time, y=Mean_Steps_By_Time)) + geom_bar(stat = "identity", fill= "springgreen4" ) +
theme(axis.text.x = element_text(angle = 90)) + labs(title="Average Total Steps vs. Time")
```

### The above visual shows a plot of average total steps against time. 

### It is noticed that activity is least from 12 AM to 4 AM. From 5 AM, the value goes on increasing almost perfectly.

### The number goes down at 3 PM, which may be assumed as lunch time for some or nap/rest time for others. 

### Bellabeat can introduce a mid day report system, for the users to know how they have fared for the past half day and what else they can do meet the acceptable quota for the remaining half of the day.




```{r}
ggplot(final, mapping=aes(x=TotalMinutesAsleep, y=SedentaryMinutes))+geom_point(color="slategray")+geom_smooth(aes(color="red"))+ labs(title="Sedantary Minutes vs. Total Minutes Asleep")
```


### This plot shows how sedentary minutes is inversely proportional to the total minutes asleep.

### In this case, Bellabeat can set an alarm for the users on their devices when their sedentary minutes have crossed a certain threshold and thus help the users focus better on getting a good sleep.

### Sedentary Minutes can be calculated for a person based on BMR (Basal metabolic Rate), which is the number of calories our body burns to stay alive. To calculate this, parameters like height, weight and age are required which are missing in the datasets.

```{r}
ggplot(final, mapping=aes(x=TotalSleepRecords,y=VeryActiveMinutes,color=TotalSleepRecords))+geom_jitter()+scale_color_gradientn(colors = c("seagreen", "lightsalmon3", "red1"))+ labs(title="Very Active Minutes vs. Total Sleep Records")
```

### Despite not being a regular direct or inverse relationship, it can be noticed here that a number of users who indulge in about one sleep session in a day have a higher value for very active minutes, while for approximately 2 sleep sessions, some users have a moderate value of very active minutes, while for three and above, the value is very low.

### As a suggestion to Bellabeat, a reminder could be set up in case the user crosses 2 sleep sessions in a day, to avoid lethargy amongst them.


```{r}
ggplot(heartandweight, mapping=aes(x=BMI,y=Value))+geom_jitter(color="red3")+geom_smooth()+ labs(title="Heart Rate vs. BMI")
```

### Unable to identify any pattern for heart rate and BMI.

### There was high scope for identifying patterns by combining weight log and heart rate data, as we could deduce multiple relationships. However, the data is minimal, and there are plenty of null values that cannot be ignored or replaced with a mean value.

### Also, combining the two datasets only provides around 10 rows which is very little data to perform further analysis. This is due to the weight log dataset being very minimal.


### Hence, more data needs to be collected on weight and mass index.


